---
title: "Geographic Dependencies within Undergradate Admissions Comeptitive Sets"
author: '@brocktibert'
date: "August 2014"
output: html_document
---

# Inspiration / Past Work

Today I came across the following post on Twitter:

> [Facebook friendships formed between people who checked in at the World Cup](https://twitter.com/seanjtaylor/status/492432734242230272/photo/1)  
  
It's not the first time that Sean has leveraged Facebook's data to explore his cool ideas: 

> [The Emotional Highs and Lows of the NFL Season](https://www.facebook.com/notes/facebook-data-science/the-emotional-highs-and-lows-of-the-nfl-season/10152033221418859)
  
  
> [NFL Fans on Facebook](https://www.facebook.com/notes/facebook-data-science/nfl-fans-on-facebook/10151298370823859)

Furthermore, Facebook has a history of highlighting (promoting?) the geo-dependency of relationships found within their platform:

> [Visualizing Friendships]("https://www.facebook.com/notes/facebook-engineering/visualizing-friendships/469716398919")


In the past, I have blogged, and presented, about the applications of [graph analysis](http://en.wikipedia.org/wiki/Graph_theory) across a wide variety of questions applicable to Enrollment Management.  

Today, I want to correlate the GPS coordinates of an insitution found in `IPEDS` (yes, this data is accessible in `IPEDS`!) to an insitution's compeitor set, which I define as the "Similar Schools" crawled from a well trafficked "College Search" site.

___more text here___




I plan on demonstrating how to crawl the web using `R` in later posts, but for time being, let's assume that we have a dataset in the following format:

```{r echo=F, comment=NA}
network[sample(1:nrow(network), size = 10, replace = F), ]
```

Here is a quick description of your database:

```{r comment=NA, echo=FALSE}
tmp_df = data.frame(column = c("from", "to", "rank"),
                    definition = c("Your Insitution's UnitID", "Your Competitor's UnitID", "   1 = Your 'biggest' competitor")
                    )
tmp_df
rm(tmp_df)

```

Because I am writing my blog post and analyzing the data simulatneously, the first column of data that you see above is an artifcat of how `R` prints data.  Basically, just ignore the first column of numbers when I print out data.  


In my experience, the `Similar Schools` feature is actually pretty accurate on a number of  __Search__ websites.  By and large, when you start to combine the data from these sites, the results actually start to mirror what I get back from the [National Student Clearinghouse](http://www.studentclearinghouse.org/).  

The key takeaway here is that these sites are pretty good lead sources, but I digress.........




## Get the Geographic Coordinates from IPEDS

For this, we are going to download a raw survey file from [IPEDS](http://nces.ed.gov/ipeds/) and read the resulting `.CSV` file into our `R` session. 



__think about breaking up the code above into descriptions__


Real quick, what do we have in our raw datasets before merging ...

-  We have `r nrow(network)` rows in the network of similar colleges   
-  The average rank is `r mean(network$rank, na.rm=T)`    
-  The number of unique schools conisdered is: `r length(unique(ids))`  


### PERHAPS I SHOULD SHOW A HISTOGRAM FOR THE RANKS



## We need to merge on the lat/long pairs for the two schools

The code below does this.  


The dataset `network` above nows shows the GPS coordinate pairs for each competitor school.





